---
title: "Covariance-Aware Augmentation Audit (E0, G0, G1, G2)"
author: "Yiming Shen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Covariance-Aware Augmentation Audit (E0, G0, G1, G2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
library(TensorEEG)
```

## What this vignette demonstrates

This vignette is the R-side counterpart of the Python protocol scripts
in `scripts/protocol/` for Shen & Degras (2026). It walks through:

1. Loading the bundled example SPD anchor stack.
2. Generating synthetic covariances with the four covariance-aware
   augmentation routines used in the multi-dataset audit ---
   E0 (off-manifold control), G0, G1, and G2.
3. Running the six-metric fidelity audit against the class-mean
   covariance reference.
4. Tabulating the results so they can be compared to the manuscript's
   Table 5.

The audit answers a single question: **for a fixed set of source
anchors, which augmentation rule keeps the synthetic covariance stack
closest to the real class geometry?** The expected ordering at the
covariance-distance metrics is

$$\text{E0} \;>\; \text{G0} \;>\; \text{G1} \;>\; \text{G2}$$

with G2 (log-Euclidean geodesic mixup) the most fidelity-preserving.

## Load the example anchors

```{r load-anchors}
data(example_anchors)
labels <- readRDS(system.file("extdata", "example_labels.rds",
                              package = "TensorEEG"))

length(example_anchors)
dim(example_anchors[[1]])
table(labels)
```

The eight anchors split evenly into class 0 (10 Hz alpha-like) and
class 1 (6 Hz lower-frequency); each is a 6 x 6 SPD covariance.

## Run the four augmentation routines

We fix the per-method seed so the table below is exactly reproducible
across machines. The augmentation ratio `n_aug = 3` matches the
revised primary protocol (synthetic:real ratio of 3 per anchor).

```{r run-augmentations}
n_aug <- 3L
g0_sigma <- 0.15

aug_g0 <- augment_cov_riemannian(
  example_anchors,
  n_aug = n_aug, sigma = g0_sigma,
  labels = labels, seed = 1002L
)

aug_e0 <- augment_cov_amplitude_matched_euclidean(
  example_anchors,
  n_aug = n_aug, g0_sigma = g0_sigma,
  labels = labels, seed = 1001L
)

aug_g1 <- augment_cov_empirical_tangent(
  example_anchors, labels,
  n_aug = n_aug, sigma = g0_sigma, seed = 1003L
)

aug_g2 <- augment_cov_geodesic_mixup(
  example_anchors, labels,
  n_aug = n_aug, beta_alpha = 1.0, seed = 1004L
)

# Each result has the same shape: 8 anchors x 3 augmentations = 24 synthetic
# covariances per method.
sapply(list(E0 = aug_e0, G0 = aug_g0, G1 = aug_g1, G2 = aug_g2),
       function(r) length(r$cov))
```

### E0 amplitude-match diagnostic

E0 is the off-manifold control: an Euclidean perturbation calibrated so
its median anchor-distance matches G0's. The match ratio rho should be
close to 1.0; a large deviation would signal that the matching routine
failed to find a target sigma. The package surfaces rho as a
diagnostic:

```{r e0-rho}
aug_e0$diagnostic
```

## Six-metric fidelity audit

`audit_covariance_fidelity()` returns the six metrics used in Table 5
of the manuscript: log-Euclidean and affine-invariant distances to the
class-mean reference, eigenvalue correlation, trace ratio, condition-
number ratio, and anchor-perturbation distance --- plus the dimension-
normalised versions of the distance metrics. We run the audit per
method and stitch the results into a single comparison table.

```{r audit}
methods <- list(E0 = aug_e0, G0 = aug_g0, G1 = aug_g1, G2 = aug_g2)

audit_one <- function(aug) {
  audit_covariance_fidelity(
    anchors    = example_anchors,
    synthetic  = aug$cov,
    anchor_idx = aug$anchor
  )
}

audits <- lapply(methods, audit_one)

audit_table <- data.frame(
  method        = names(audits),
  LE_distance   = sapply(audits, `[[`, "log_euclidean_to_reference"),
  AI_distance   = sapply(audits, `[[`, "affine_invariant_to_reference"),
  trace_ratio   = sapply(audits, `[[`, "trace_ratio"),
  cond_ratio    = sapply(audits, `[[`, "condition_number_ratio"),
  anchor_dist   = sapply(audits, `[[`, "anchor_perturbation_distance"),
  LE_normalized = sapply(audits, function(a) a$normalized$log_euclidean),
  row.names     = NULL,
  stringsAsFactors = FALSE
)
audit_table[, -1] <- round(audit_table[, -1], 4)
audit_table
```

### How to read the table

- **LE_distance / AI_distance.** Lower is better: synthetic stack is
  closer in covariance geometry to the class-mean reference. The
  expected ordering is E0 > G0 > G1 > G2.
- **trace_ratio.** Closer to 1.0 means the synthetic mean preserves the
  total power of the reference. Note that G0 can inflate this even when
  it preserves SPD-manifold membership.
- **cond_ratio.** Closer to 1.0 means the synthetic mean preserves
  conditioning. G0 isotropic jitter can degrade conditioning relative
  to E0 even though E0 is off-manifold.
- **anchor_dist.** Mean Frobenius distance from each synthetic
  covariance back to its anchor in log-Euclidean coordinates.
- **LE_normalized.** Distance divided by sqrt(p(p+1)/2). This is the
  metric used for cross-channel-count pooling in Table 5 of the
  manuscript.

This is the same multi-metric profile reported in the paper: distance
metrics rank E0/G0/G1/G2 monotonically, but trace and conditioning
metrics show that G0's isotropic jitter has its own fidelity costs.
That is why the manuscript treats fidelity as a multi-metric audit
rather than a single pass/fail score.

## Visualisation

A bar chart of the normalised LE distance makes the ranking
unambiguous:

```{r plot, fig.width = 6, fig.height = 4}
op <- par(mar = c(4, 4.2, 3, 1))
barplot(
  audit_table$LE_normalized,
  names.arg = audit_table$method,
  col = c("#d95f02", "#7570b3", "#1b9e77", "#66c2a5"),
  main = "Dimension-normalised LE distance to class mean",
  ylab = "d_LE / sqrt(p(p+1)/2)",
  border = NA
)
par(op)
```

## What is and is not in the audit

This vignette runs the audit on a small two-class toy stack so it
finishes in seconds. The full multi-dataset audit in the manuscript:

- Pools fidelity metrics across BNCI2014_001, BNCI2014_004, Zhou2016
  and Stieger2021 (N = 32 subjects, p ranging from 3 to 60).
- Uses the calibration manifest written by the Python driver to fix
  the random anchors-per-resample exactly.
- Reports paired Wilcoxon tests of E0 vs each geometry-preserving
  variant.

To use the manifest-based replay yourself, see
`vignette("manifest-replay")`.

## Reference

Shen, Y., & Degras, D. (2026). *Covariance Geometry as a Safety
Constraint for Cross-Session BCI Augmentation: A Multi-Dataset
Non-Inferiority and Fidelity Audit.*
