---
title: "Manifest-Based Replay (Cross-Language Reproducibility)"
author: "Yiming Shen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Manifest-Based Replay (Cross-Language Reproducibility)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(TensorEEG)
```

## Why a manifest?

The audit in Shen & Degras (2026) writes a `calibration_manifest.csv`
for every `(dataset, subject, budget, resample, method, class)` cell.
Each row records:

- the real source-session trial ids that served as anchors,
- the per-method seed used to draw the synthetic stack,
- the augmentation ratio,
- a `target_session_ids_used` flag for the leakage audit.

This manifest is the audit trail that lets a third party regenerate
the synthetic covariance stack exactly --- without rerunning the
Python protocol driver or accessing any private intermediate state.
The manifest is the contract; the seeds are the keys.

The `read_calibration_manifest()` and `replay_from_manifest()`
functions are the R-side analogues of
`scripts/protocol/manifest.py` and
`scripts/protocol/run_fidelity.py::_regenerate_synthetic`. Together
they let an R user reproduce the per-cell synthetic stack and rerun
the fidelity audit without leaving R.

## Demo manifest shipped with the package

A tiny manifest is bundled in `inst/extdata/`. It describes a single
(`Example`, subject 1, budget 4, resample 0) cell with five method
codes (R0/E0/G0/G1/G2) and two classes per method. Eight `Example`
anchors are bundled separately as `data(example_anchors)`.

```{r read-manifest}
manifest_path <- system.file("extdata", "example_manifest.csv",
                             package = "TensorEEG")
manifest <- read_calibration_manifest(manifest_path)

str(manifest, max.level = 1)
head(manifest[, c("method_code", "class_label", "n_real_trials",
                  "method_seed", "n_synthetic_per_anchor", "notes")])
```

`read_calibration_manifest()` parses the JSON-like `real_trial_ids`
column into a list-column of integer vectors:

```{r real-ids}
manifest$real_trial_ids[1:2]
```

These are 0-based to follow the Python protocol convention.
`replay_from_manifest()` converts them to 1-based R indices when it
selects anchors from `source_cov_list`.

## Replay synthetic stacks for every method

```{r load-anchors}
data(example_anchors)
anchor_labels <- readRDS(system.file("extdata", "example_labels.rds",
                                     package = "TensorEEG"))
length(example_anchors)
```

We replay the cell for each method code, holding the anchors and
labels constant:

```{r replay-each}
methods_to_replay <- c("R0", "E0", "G0", "G1", "G2")

replays <- lapply(methods_to_replay, function(m) {
  group <- manifest[manifest$method_code == m, , drop = FALSE]
  replay_from_manifest(
    manifest_group  = group,
    source_cov_list = example_anchors,
    source_labels   = anchor_labels,
    ratio_per_anchor = 3L,
    g0_sigma = 0.15,
    g1_sigma = 0.15,
    g2_beta_alpha = 1.0
  )
})
names(replays) <- methods_to_replay

sapply(replays, function(r) length(r$cov))
```

R0 returns an empty stack by design (real-only baseline; no synthetic
covariances are generated). The other four methods return
`8 anchors x 3 augmentations = 24` synthetic covariances each.

## Verify byte-equivalent reproducibility

A single seed-driven call to `augment_cov_geodesic_mixup()` directly
should produce a synthetic stack identical to the manifest-driven
replay --- because the manifest replay forwards the same seed to the
same routine. This is the integration test we report in Section 6 of
the manuscript.

```{r byte-equiv}
direct_g2 <- augment_cov_geodesic_mixup(
  cov_list   = example_anchors,
  labels     = anchor_labels,
  n_aug      = 3L,
  beta_alpha = 1.0,
  seed       = 1004L  # matches example_manifest.csv method_seed for G2
)

stopifnot(length(direct_g2$cov) == length(replays$G2$cov))

max_abs_diff <- max(mapply(function(A, B) max(abs(A - B)),
                           direct_g2$cov, replays$G2$cov))
max_abs_diff
```

The maximum absolute difference should be exactly 0 --- not merely
floating-point close. Equivalent guarantees hold for E0, G0, and G1.

## Run the fidelity audit on the replayed stacks

The audit machinery is method-agnostic, so we can run it directly on
the replayed synthetics:

```{r audit-replays}
audit_replays <- function(rep_list) {
  if (length(rep_list$cov) == 0L) {
    return(c(LE = NA_real_, AI = NA_real_, anchor = NA_real_))
  }
  res <- audit_covariance_fidelity(
    anchors    = example_anchors,
    synthetic  = rep_list$cov,
    anchor_idx = rep_list$anchor
  )
  c(LE     = res$log_euclidean_to_reference,
    AI     = res$affine_invariant_to_reference,
    anchor = res$anchor_perturbation_distance)
}

audit_table <- t(sapply(replays, audit_replays))
round(audit_table, 4)
```

The ranking matches `vignette("covariance-audit")`, but every number
here was reproduced from a CSV file plus a seed --- the kind of audit
trail a reviewer can verify without trusting the original run.

## Using a manifest from your own protocol run

If you have run the Python driver yourself, just point
`read_calibration_manifest()` at the manifest CSV it wrote:

```r
manifest <- read_calibration_manifest(
  "results/protocol/sanity_B100/calibration_manifest.csv"
)
```

You will then need:

- `source_cov_list`: SPD covariances for all source-session trials
  (1-based indexing in R; the manifest's 0-based ids are translated
  internally).
- `source_labels`: integer labels matching `source_cov_list`.

`replay_from_manifest()` validates that each cell contains exactly one
row per class; if a cell is missing or malformed it stops with an
explicit message rather than silently producing the wrong stack.

## Note on A0 (alignment)

A0 is intentionally **not** replayable from a manifest alone, because
Riemannian alignment is transductive --- it requires the unlabeled
target-session covariance stack at fit time, not just a seed. Use
`augment_cov_alignment_riemannian()` directly with the target stack
when reproducing A0 results.

## Reference

Shen, Y., & Degras, D. (2026). *Covariance Geometry as a Safety
Constraint for Cross-Session BCI Augmentation: A Multi-Dataset
Non-Inferiority and Fidelity Audit.*
