---
title: "TensorEEG Validation: Spatial & Temporal Covariance"
author: "Yiming Shen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{TensorEEG Validation: Spatial & Temporal Covariance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
library(TensorEEG)
```

## Purpose

This vignette is the validation walkthrough for the TensorEEG simulation
engine. It mathematically verifies that the synthetic data produced by
`sim_eeg_master()` adheres to two physical constraints of real EEG:

1. **Spatial domain (volume conduction & rank deficiency).** Real EEG is
   not full rank --- a small number of latent cortical sources drive
   many sensors. We check this with the eigenvalue spectrum of the
   spatial covariance and look for off-diagonal sensor coupling
   indicative of volume conduction.

2. **Temporal domain (oscillatory dynamics).** The VAR(2) background
   process should generate stable oscillations rather than white noise.
   We verify this with the temporal covariance structure and the
   autocorrelation function (ACF), which should peak at lags
   corresponding to the configured target frequencies.

Successful validation confirms the synthetic tensor is suitable for
tensor decomposition, Riemannian-geometry benchmarking, or as the
SimEEG positive-control input to the covariance-aware augmentation
audit (see `vignette("covariance-audit")`).

## Generate a synthetic tensor

```{r run-sim}
set.seed(42)
fixed_freqs <- c(10, 12, 10, 10, 20, 20, 22, 20, 10, 20)
sim_result <- sim_eeg_master(
  n_trials = 10,
  n_time = 500,
  n_channels = 64,
  target_freqs = fixed_freqs,
  drift_power_ratio = 0.5
)
str(sim_result, max.level = 2)
```

We extract the first trial and apply a 1 Hz high-pass filter to remove
slow drift. Slow drift is real but would mask the fine covariance
structure we want to validate here.

```{r prep}
X_raw <- sim_result$data[, , 1]
fs <- sim_result$params$fs
bf_hp <- signal::butter(4, 1 / (fs / 2), type = "high")
X_neural <- apply(X_raw, 2, function(x) signal::filtfilt(bf_hp, x))
dim(X_neural)
```

## Module A: spatial covariance

Objective: verify volume conduction effects and low-rank structure.

```{r spatial}
S_cov <- cov(X_neural)
eigen_S <- eigen(S_cov, only.values = TRUE)$values
eigen_norm <- eigen_S / sum(eigen_S)
round(head(eigen_norm, 12), 4)
```

The first ten eigenvalues should dominate the spectrum, reflecting the
ten latent sources used to generate the trial.

## Module B: temporal covariance and oscillatory content

Objective: verify oscillatory dynamics and stationarity. We compute the
temporal covariance over the first 0.8 s and the average ACF across
sensors, looking for peaks at the lags corresponding to 10 Hz (100 ms)
and 20 Hz (50 ms).

```{r temporal}
t_subset <- 1:200
X_sub <- X_neural[t_subset, ]
X_sub <- sweep(X_sub, 2, colMeans(X_sub), "-")
T_cov <- (X_sub %*% t(X_sub)) / ncol(X_sub)
dim(T_cov)

acf_avg <- rowMeans(apply(X_neural, 2, function(x) {
  acf(x, lag.max = 100, plot = FALSE)$acf
}))
lags_ms <- (0:100) / fs * 1000
```

## Visualisation dashboard

```{r dashboard, fig.width = 8, fig.height = 7}
op <- par(mfrow = c(2, 2), mar = c(4, 4, 3, 2))

# Plot 1: spatial covariance heatmap
image(S_cov[, ncol(S_cov):1],
      main = "Spatial covariance (64 x 64)",
      axes = FALSE,
      col = grDevices::hcl.colors(20, "YlOrRd", rev = TRUE))
mtext("Sensor i", side = 2, line = 1)
mtext("Sensor j", side = 1, line = 1)
box()

# Plot 2: scree plot
plot(eigen_norm[1:20], type = "b", pch = 19, col = "blue", lwd = 2,
     main = "Spatial eigenvalues (rank test)",
     xlab = "Component index", ylab = "Explained variance ratio")
abline(v = 10, col = "red", lty = 2)
text(10, max(eigen_norm) * 0.8, "True sources = 10",
     col = "red", pos = 4)

# Plot 3: temporal covariance
image(T_cov[, ncol(T_cov):1],
      main = "Temporal covariance (0-0.8 s)",
      axes = FALSE,
      col = grDevices::hcl.colors(20, "Viridis"))
mtext("Time t1", side = 2, line = 1)
mtext("Time t2", side = 1, line = 1)
box()

# Plot 4: average ACF
plot(lags_ms, acf_avg, type = "l", lwd = 2, col = "darkgreen",
     main = "Average autocorrelation (ACF)",
     xlab = "Lag (ms)", ylab = "Correlation")
abline(h = 0, col = "gray", lty = 2)
abline(v = c(50, 100), col = "red", lty = 3)
text(50,  0.5, "20 Hz",  col = "red", pos = 3, cex = 0.8)
text(100, 0.5, "10 Hz",  col = "red", pos = 3, cex = 0.8)

par(op)
```

### Interpretation

- **Spatial covariance heatmap.** Block-like off-diagonal structure
  reflects neighbouring-sensor coupling from volume conduction.
- **Scree plot.** Eigenvalues should drop sharply after index 10 and
  enter a noise plateau, confirming the latent-source rank.
- **Temporal covariance.** A checkerboard pattern parallel to the main
  diagonal indicates periodic oscillatory structure rather than
  unstructured white noise.
- **Average ACF.** Distinct side-lobe peaks at 50 ms and 100 ms
  correspond to the 20 Hz beta and 10 Hz alpha frequencies configured
  in `target_freqs`.

If all four panels look as described, the simulation engine is producing
trials whose spatial-temporal covariance structure matches the
neurophysiological priors used to design TensorEEG.

## Next steps

- See `vignette("covariance-audit")` for an end-to-end demonstration of
  the four covariance augmentation routines (E0, G0, G1, G2) and the
  six-metric fidelity audit on these covariances.
- See `vignette("manifest-replay")` for using the calibration manifest
  written by the Python protocol driver to replay synthetic stacks
  byte-equivalently in R.

## Reference

Shen, Y., & Degras, D. (2026). *Covariance Geometry as a Safety
Constraint for Cross-Session BCI Augmentation: A Multi-Dataset
Non-Inferiority and Fidelity Audit.*
